The Impact of News Sentiment on MSFT

### 2023-12-21 MSFT.Rmd
hopefully final report

library(PerformanceAnalytics)
library(forecast)
library(rugarch)
library(vars)
library(tseries)
library(quantmod)
library(tinytex)
library(zoo)
library(ggplot2)
library(utils)
library(dplyr)
library(xts)
library(timeSeries)
```

@@

MSFT_VAR <- VAR(MSFT_VAR_Data, ic="AIC", lag.max = 24)
coeftest(MSFT_VAR)
```
We see that almost none of the coefficients have a statistical significance as their p-values are all above 0.05. An exception is the second lag of MSFT returns. It seems to have a slight negative effect on the current price of MSFT. However, there is only small evidence.

@@

## Forecasting
From a practical point of view, our VAR model may still be of some use when applied for forecasting purposes. To see if the inclusion of the news sentiment in our model has an improving effect on the models accuracy when predicting the price movements of MSFT we do some forecasting for our testing period where we include the news sentiment data of the testing period.

@@

```{r echo=FALSE, include=TRUE}
forecast_data <- data.frame(
  MSFT_Returns = rep(NA, length(news_test)),
  News_Sentiment = news_test
)

MSFT_VAR_pred <- predict(MSFT_VAR, newdata = forecast_data, n.ahead=43)
plot(MSFT_VAR_pred)
MSFT_VAR_pred_values <- MSFT_VAR_pred$fcst$MSFT_Returns

rmse_VAR <- sqrt(mean((MSFT_returns_test - MSFT_VAR_pred_values)^2))
cat("RMSE of VAR\t\t\t", rmse_VAR)
```

```{r echo=FALSE, include=TRUE}
MSFT_ARMA <- auto.arima(MSFT_returns_train)
MSFT_ARMA
MSFT_ARMA_pred <- forecast(MSFT_ARMA, level=95, h = length(news_test))
plot(MSFT_ARMA_pred)
```

```{r echo=FALSE, include=TRUE}
rmse_ARMA <- sqrt(mean((MSFT_returns_test - MSFT_ARMA_pred$mean)^2))
cat("RMSE of AR(2) Benchmark \t", rmse_ARMA)
```
We see that auto.arima has found that the order of the auto-regressive (AR) as well as the Moving Average (MA) parts are of order 0. Therefore, we essentially deal with a mean model as our benchmark. During forecasting, we also see that it uses a mean of 0 with a confidence band for all future values. This makes sense, as we deal with stationary data that commonly has a mean of 0.

@@

In summary, our comprehensive analysis using an ARIMA model with Fourier terms suggests that while there is a discernible pattern in the data, it is not strongly influenced by news sentiment as originally hypothesized. The non-significant results of the Granger causality test, coupled with the model's performance metrics, indicate that Microsoft's stock price movements are likely driven by factors not captured by news sentiment from the selected subreddits. The model's predictive performance indicates an area for potential improvement over a simpler benchmark model.

@@

>>>>

library(PerformanceAnalytics)
library(forecast)
library(rugarch)
library(vars)
library(urca)
library(tseries)
library(quantmod)
library(tinytex)
library(zoo)
library(ggplot2)
library(utils)
library(dplyr)
library(xts)
library(timeSeries)
```

@@

# lag.max = 24 is a starting point for high-frequency data (like daily data), 
# we consider a larger maximum lag because there could be significant autocorrelations over several days.
MSFT_VAR <- VAR(MSFT_VAR_Data, ic="AIC", lag.max = 24)
coeftest(MSFT_VAR)
```
We see that almost none of the coefficients have a statistical significance as their p-values are all above 0.05. An exception is the second lag of MSFT returns. It seems to have a slight negative effect on the current price of MSFT. However, there is only small evidence.

We tried another method of sentiment analysis, VADER (Valence Aware Dictionary and Sentiment Reasoner) is a lexicon and rule-based sentiment analysis tool that assigns sentiment scores to words and combines them to determine the overall sentiment polarity and intensity of a piece of text, with a focus on handling sentiment expressions in informal language such as social media:

```{r echo=FALSE, include=TRUE}
MSFT.vadar <- read.csv("data/final_data_vadar.csv")
MSFT.vadar$Index <- as.Date(MSFT.vadar$Index)
MSFT.vadar$MSFT.Adjusted <- as.numeric(as.character(MSFT.vadar$MSFT.Adjusted))
MSFT.vadar$sentiment_average_score <- as.numeric(as.character(MSFT.vadar$sentiment_average_score))
MSFT.vadar_ts <- zoo(MSFT.vadar$MSFT.Adjusted, order.by = MSFT.vadar$Index)
MSFT.vadar_returns <- na.omit(diff(log(MSFT.vadar_ts)))

vadar.news_ts <- zoo(MSFT.vadar$sentiment_average_score, order.by = MSFT.vadar$Index)

MSFT.vadar_returns_train <- window(MSFT.vadar_returns, start ="2022-01-04", end="2023-09-30")
MSFT.vadar_returns_test <- window(MSFT.vadar_returns, start="2023-10-01")

vadar.news_train <- window(vadar.news_ts, start ="2022-01-04", end="2023-09-30")
vadar.news_test <- window(vadar.news_ts, start="2023-10-01")

MSFT.vadar_VAR_Data <- cbind(MSFT.vadar_returns_train, vadar.news_train)
colnames(MSFT.vadar_VAR_Data) <- c("MSFT_Returns", "News_Sentiment")

MSFT.vadar_VAR <- VAR(MSFT.vadar_VAR_Data, ic="AIC", lag.max = 24)
coeftest(MSFT.vadar_VAR)
```
Overall, this VAR model on the new sentiment analysis indicates that certain past values Microsoft stock returns and reddit posts have a significant impact on the current values of these variables. The model suggests a complex interplay between stock returns and reddit posts sentiment with specific lags being more influential than others. However, many of the coefficients are not significant, suggesting that not all past values are equally informative for predicting current values.

@@

```{r echo=FALSE, include=TRUE}
# Casuality test for VADAR
causality(MSFT.vadar_VAR, cause="News_Sentiment")["Granger"] 
```
The p-value of 0.008237 from the Granger causality test suggests that reddit posts sentiment is indeed a causal driver of MSFT stock price. This result leads us to accept the hypothesis that changes in news sentiment have a predictive effect on Microsoft stock returns.

## Forecasting
We will use the VADAR VAR model for forecasting purposes to see if the inclusion of the reddit posts sentiment in our model has an improving effect on the models accuracy when predicting the price movements of MSFT we do some forecasting for our testing period where we include the reddit posts sentiment data of the testing period.

@@

```{r echo=FALSE, include=TRUE}
forecast_data <- data.frame(
  MSFT_Returns = rep(NA, length(vadar.news_test)),
  News_Sentiment = vadar.news_test
)

MSFT_VAR_pred <- predict(MSFT.vadar_VAR, newdata = forecast_data, n.ahead=43)
plot(MSFT_VAR_pred)
MSFT_VAR_pred_values <- MSFT_VAR_pred$fcst$MSFT_Returns

rmse_VAR <- sqrt(mean((MSFT.vadar_returns_test - MSFT_VAR_pred_values)^2))
cat("RMSE of VAR\t\t\t", rmse_VAR)
```

```{r echo=FALSE, include=TRUE}
MSFT_ARMA <- auto.arima(MSFT.vadar_returns_train)
MSFT_ARMA
MSFT_ARMA_pred <- forecast(MSFT_ARMA, level=95, h = length(vadar.news_test))
plot(MSFT_ARMA_pred)
```

```{r echo=FALSE, include=TRUE}
rmse_ARMA <- sqrt(mean((MSFT.vadar_returns_test - MSFT_ARMA_pred$mean)^2))
cat("RMSE of AR(2) Benchmark \t", rmse_ARMA)
```
We see that auto.arima has found that the order of the auto-regressive (AR) as well as the Moving Average (MA) parts are of order 0. Therefore, we essentially deal with a mean model as our benchmark. During forecasting, we also see that it uses a mean of 0 with a confidence band for all future values. This makes sense, as we deal with stationary data that commonly has a mean of 0.

@@

In summary, our comprehensive analysis using an ARIMA model with the addition of VADER sentiment analysis suggests offers new insights into the influence of reddit posts sentiment on Microsoft's stock price. The updated Granger causality test results from the VADER model indicate a possible predictive effect of subreddit sentiment on stock returns, a finding that warrants further exploration. While the earlier model did not show a strong sentiment impact, the VADER analysis suggests that sentiment may play a role, although it may not be the primary driver of stock movements. The significant results of the Granger causality test, coupled with the model's performance metrics, indicate that Microsoft's stock price movements are likely driven by factors not captured by news sentiment from the selected subreddits. The model's predictive performance indicates an area for potential improvement over a simpler benchmark model.
In addition, our examination of seasonality using the Fourier transform revealed complex seasonal patterns in the data. This suggests that stock price movements may be influenced by a combination of factors, including subtle seasonal trends and shifts in sentiment. Our findings suggest potential areas for improvement in modeling and forecasting, and underscore the need for models that can capture these multifaceted influences.

@@

"The Causal Relationship between Social Media Sentiment and Stock Return: Experimental Evidence from an Online Message Forum" by Lin et al. (2022) examines the causal effect of social media sentiment on stock returns using an experimental approach. The study finds that positive sentiment can significantly increase same-day stock prices, but the effect is short-lived.

"Impact Of Social Media On Stock Market- A Case Of Sentiment Analysis Using Vader" by Kumar et al. (2021) applies the Vader sentiment analysis tool to Twitter data to assess its impact on stock market movements. The study finds that positive sentiment tweets are associated with increased stock prices, while negative sentiment tweets are associated with decreased stock prices.

### 2023-12-21 MSFT_with_vadar_analysis/MSFT_vadar_analysis.Rmd
hopefully final report

```{r plot_chunk, fig.height=10, fig.width=14, echo=FALSE, include=TRUE}
MSFT <- read.csv("final_data_vadar.csv")
MSFT$Index <- as.Date(MSFT$Index)
MSFT$MSFT.Adjusted <- as.numeric(as.character(MSFT$MSFT.Adjusted))
MSFT$sentiment_average_score <- as.numeric(as.character(MSFT$sentiment_average_score))
MSFT_ts <- zoo(MSFT$MSFT.Adjusted, order.by = MSFT$Index)

events <- data.frame(
  date = as.Date(c("2023-01-23", "2022-11-30", "2023-06-29", "2023-04-12", "2023-11-15")),
  event = c(
    "Investing into OpenAI",
    "ChatGPT",
    "Bing Co-Pilot Release",
    "DALL-E 3 Release",
    "Sam Altman Drama"
  )
)

event_data <- merge(events, MSFT, by.x = "date", by.y = "Index", all.x = TRUE)

ix.na = is.na(MSFT$sentiment_average_score)
problematic_values <- MSFT$sentiment_average_score[ix.na]
MSFT$sentiment_average_score[problematic_values] <- NA
# Replace with 0
MSFT$sentiment_average_score[problematic_values] <- 0
MSFT$Index_numeric <- as.numeric(MSFT$Index - as.Date("2000-01-01"))
MSFT$smoothed_polarity_loess <- with(MSFT, loess(sentiment_average_score ~ Index_numeric, span = 0.2)$fitted)

main_plot <- ggplot(data = MSFT, aes(x = Index)) +
  geom_line(aes(y = MSFT.Adjusted, color = "MSFT.Adjusted"), size = 1) +
  ylab("MSFT Adjusted Price") +

  geom_line(aes(y = sentiment_average_score, color = "Sentiment Average Score"), size = 1) +
  ylab("Sentiment Average Score") +

  scale_y_continuous(
    name = "MSFT Adjusted Price",
    sec.axis = sec_axis(~./2000, name = "Sentiment Average Score"),
    limits = c(0, 400),
    breaks = seq(0, 400, by = 50)
  ) +

  scale_color_manual(
    values = c("MSFT.Adjusted" = "#00AFBB", "Sentiment Average Score" = "red"),
    labels = c("MSFT Adjusted Price", "Sentiment Average Score")
  ) +

  ggtitle("MSFT Adjusted Price and Sentiment Average Score Over Time") +
  theme_minimal()

# Adjust the legend size and position
main_plot <- main_plot + theme(legend.position = c(0.15, 0.2), 
                               legend.background = element_blank(), # Transparent background
                               legend.key = element_blank(), 
                               legend.text = element_text(size = 8)) # Smaller text size

# arrows and text layer
arrows_layer <- ggplot(event_data, aes(x = date, y = MSFT.Adjusted, label = event)) +
  geom_segment(aes(xend = date, yend = MSFT.Adjusted - 20), 
               arrow = arrow(type = "closed", length = unit(0.1, "inches")),
               size = 0.5) +
  labs(x = NULL, y = NULL) +  # Remove axis labels
  geom_text(hjust = 0.6, vjust = -0.7, size = 2.5, color = "black") +  # Add text
  labs(x = NULL, y = NULL) +  
  theme_void()

combined_plot <- main_plot + annotation_custom(
  ggplotGrob(arrows_layer),
  xmin = min(event_data$date), xmax = max(event_data$date),
  ymin = min(event_data$MSFT.Adjusted) + 10, ymax = max(event_data$MSFT.Adjusted) + 10
)

print(combined_plot)
```
We see a clear trend, which is an indicator for non-stationarity, so we do an Augmented Dickey-Fuller (ADF) Test to see whether the data is in fact non-stationary.

>>>>

```{r plot_chunk, fig.height=10, fig.width=14, echo=FALSE, include=TRUE}
MSFT <- read.csv("final_data_vadar.csv")
# MSFT$Index <- as.Date(MSFT$Index)
MSFT$MSFT.Adjusted <- as.numeric(as.character(MSFT$MSFT.Adjusted))
MSFT_ts <- zoo(MSFT$MSFT.Adjusted, order.by = MSFT$Index)

MSFT$Index <- as.Date(MSFT$Index)
ratio <- max(MSFT$MSFT.Adjusted, na.rm = TRUE) / max(MSFT$sentiment_average_score, na.rm = TRUE)

# Create a basic plot
p <- ggplot(MSFT, aes(x = Index)) +
  geom_line(aes(y = MSFT.Adjusted, colour = "MSFT Adjusted")) +
  geom_line(aes(y = sentiment_average_score, colour = "Sentiment Average Score")) +
  scale_colour_manual("", 
                      breaks = c("MSFT Adjusted", "Sentiment Average Score"),
                      values = c("blue", "red")) +
  scale_y_continuous(
    "MSFT Adjusted",
    sec.axis = sec_axis(~ . / ratio, name = "Sentiment Average Score")
  ) +
  labs(title = "MSFT Adjusted and Sentiment Average Score Over Time",
       x = "Date",
       y = "MSFT Adjusted")

# Show the plot
print(p)
```
We see a clear trend, which is an indicator for non-stationarity, so we do an Augmented Dickey-Fuller (ADF) Test to see whether the data is in fact non-stationary.
